---
title: "Intergenic"
author: "Szi Kay Leung"
date: "06/09/2023"
output:
  rmdformats::robobook:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Aim

With over 40% of transcripts in the whole transciptome dataset annotated as "Intergenic", this rmd aims to investigate further the features of these transcripts.

```{r input}
suppressMessages(library("data.table"))
suppressMessages(library("ggplot2"))
suppressMessages(library("stringr"))
suppressMessages(library("dplyr"))

LOGEN <- "/gpfs/mrc0/projects/Research_Project-MRC148213/sl693/scripts/LOGen/"
source(paste0(LOGEN,"transcriptome_stats/read_sq_classification.R"))

input.file <- "/gpfs/mrc0/projects/Research_Project-MRC148213/Rosie/WholeTargeted/SQANTI/WholeTargeted_cleaned_aligned_merged_collapsed_qced_RulesFilter_2reads2samples_classification.txt"
input.file <- SQANTI_class_preparation(input.file,"nstandard")

```

## Structural categories

SQANTI annotates transcripts using the following categories:  
`r unique(input.file$structural_category)`.  

Of note, transcripts that are between genes but are antisense to a known gene with potential regulatory functions are classified as "antisense". Conversely, transcripts that are between genes are classified as "intergenic".

## Exploration {.tabset .tabset-fade}

```{r intergenic}
Intergenic <- input.file[input.file$structural_category == "Intergenic",1:10]
monoIntergenic <- Intergenic[Intergenic$exons == 1,]
```

### Mono-exons
Vast majority of intergenic are mono-exonic: `r nrow(monoIntergenic)` (`r round(nrow(monoIntergenic)/nrow(Intergenic) * 100,2)`%)

```{r intergenic1}
Intergenic %>% group_by(exons) %>% tally() %>% ggplot(.,aes(x=exons,y=n)) + geom_bar(stat="identity") +
  labs(x = "Exon number", y = "Number of transcripts")
```

### Mono-exons by length

Of these mono-exonic intergenic transcripts, two-thirds are less than 1kb: `r nrow(Intergenic[Intergenic$length < 1000,])` (`r round(nrow(Intergenic[Intergenic$length < 1000,])/nrow(monoIntergenic) * 100,2)`%)

```{r intergenic2}
ggplot(Intergenic,aes(length)) + geom_histogram(bins=40) + labs(x = "Length (bp)", y = "Number of transcripts")
```

### Distribution aross genome

The distribution of these mono-exonic transcripts are not uniform across the genome, with notably enrichment in chromosome 1 and 2.

```{r}
Intergenic %>% mutate(chrom = str_remove(chrom,"chr")) %>% group_by(chrom) %>% tally() %>% ggplot(.,aes(x=reorder(chrom,-n),y=n)) + geom_bar(stat="identity") + labs(x = "Chromosome", y = "Number of transcripts")
```

